\vspace{-12pt}\large \textbf{Burrows-Wheeler transform for terabases}

### 2016-01-05 paper/paper.tex
Changes to the final DCC paper.

\author{%
Jouni Sirén\thanks{This work was supported by the Wellcome Trust grant [098051].} \\[0.5em]
{\small\begin{minipage}{\linewidth}\begin{center}
\begin{tabular}{c}
Wellcome Trust Sanger Institute \\
Wellcome Trust Genome Campus \\
Hinxton, Cambridge, CB10 1SA, UK \\
\url{jouni.siren@iki.fi}
\end{tabular}
\end{center}\end{minipage}}
}

@@

The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and to derive the \BWT{} from the suffix array. These algorithms cannot be used with large datasets, as they require much more memory than the sequences themselves. Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
% DCC:
\cite{Kaerkkaeinen2015a}.
% arXiv:
%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.

There are many \emph{direct} \BWT{} construction algorithms that do not need the suffix array. Some require a limited amount of working space on top of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, while others use the disk as additional working space \cite{Ferragina2012,Beller2013}. These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s. \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs. Some can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming the bottleneck after about 1~Tbp \cite{Bauer2013,Li2014a}. GPU-based algorithms are even faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}. Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, at the price of using more resources for the construction \cite{Wang2015}.

@@

When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling. Because all endmarkers have distinct values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the end of text $T_{i}$. To find the end, we iterate $\Psi(i) = \mBWT.\mselect(i - \mC[c], c)$, where $c$ is the largest value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$. If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$, for $1 \le j < \abs{T_{i}}$. As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.

@@

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{}.
% arXiv:
%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
%
We can reduce the overhead by writing the lexicographic ranks to \emph{disk}. If we sort the ranks on disk, we just need to scan the rank array once during the merge phase. We can also \emph{compress} the ranks before writing them to disk and \emph{interleave} the sorting with the search and merge phases. We now describe the key ideas for fast and space-efficient \BWT{}.

@@

Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$. As suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, we can output a \emph{run} of ranks $(r, sp+1-ep)$. Afterwards, we proceed to the \emph{children} of the node. For each character $c \in \Sigma$, we create a node corresponding to suffix $cX$, rank $\mLF_{\Acoll}(r,c)$, and range $[\mLF_{\Bcoll}(sp-1, c) + 1, \mLF_{\Bcoll}(ep, c)]$. Searching the branches of the trie can be done in parallel using multiple \emph{threads}.

>>>>

\author{%
Jouni Sirén\thanks{This work was supported by the Wellcome Trust grant [098051].} \\[0.5em]
{\small\begin{minipage}{\linewidth}\begin{center}
\begin{tabular}{c}
Wellcome Trust Sanger Institute \\
Wellcome Genome Campus \\
Hinxton, Cambridge, CB10 1SA, UK \\
\url{jouni.siren@iki.fi}
\end{tabular}
\end{center}\end{minipage}}
}

@@

The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and then derive the \BWT{} from the suffix array. These algorithms cannot be used with large datasets, as they require much more memory than the sequences themselves. Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
% DCC:
\cite{Kaerkkaeinen2015a}.
% arXiv:
%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.

There are many \emph{direct} \BWT{} construction algorithms that do not need the suffix array. Some require a limited amount of working space on top of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, while others use the disk as additional working space \cite{Ferragina2012,Beller2013}. These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s. \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs. Some can index 5\nobreakdash--10~Mbp/s using ordinary hardware, with their memory usage becoming the bottleneck after about 1~Tbp \cite{Bauer2013,Li2014a}. GPU-based algorithms are even faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}. Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, at the price of using more resources for the construction \cite{Wang2015}.

@@

When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling. Because all endmarkers have distinct values during sorting, we know that $\mSA[i]$ with $i \le m$ points to the end of text $T_{i}$. To find the end, we iterate $\Psi(i) = \mBWT.\mselect(i - \mC[c], c)$, where $c$ is the largest value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$. If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$, for $1 \le j \le \abs{T_{i}}$. As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.

@@

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{}.
% arXiv:
%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
%
We can reduce the overhead by writing the lexicographic ranks to \emph{disk}. If we sort the ranks on disk, we just need to scan the rank array once during the merge phase. We can also \emph{compress} the ranks before writing them to disk and \emph{interleave} the sorting with the search and merge phases. We now describe the key ideas for fast and space-efficient \BWT{} construction.

@@

Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$. As suffix $X$ occurs $ep+1-sp$ times in collection $\Bcoll$, we can output a \emph{run} of ranks $(r, ep+1-sp)$. Afterwards, we proceed to the \emph{children} of the node. For each character $c \in \Sigma$, we create a node corresponding to suffix $cX$, rank $\mLF_{\Acoll}(r,c)$, and range $[\mLF_{\Bcoll}(sp-1, c) + 1, \mLF_{\Bcoll}(ep, c)]$. Searching the branches of the trie can be done in parallel using multiple \emph{threads}.
